---
title: "Sampling variance in MIAMA counterfactual assessments"
subtitle: "ARCHIVED: pre-canonical-assumptions, pre-T3 implementation"
author: "MIAMA project team"
date: last-modified
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    embed-resources: true
  pdf:
    toc: true
    number-sections: true
    geometry:
      - margin=22mm
params:
  dataset_size: "leeds"
  geography_id: "E08000035"
  assessment_period_years: 10
  n_replicates: 10
  effect_percent: "10,20,30,40,50,60,70,80,90,100"
  include_no_change: true
  sd_bootstrap_draws: 1000
  run_sample_size_experiment: false
  sample_sizes: "250,500,1000,2500,5000"
  sample_size_effect_percent: "10,50,100"
  sample_size_replicates: 10
  run_combined_mode_experiment: false
  combined_effect_percent: "10,50,100"
  combined_replicates: 10
  master_seed: 20260827
  run_simulation: true
  reuse_cached_runs: true
  cache_version: "v3"
execute:
  echo: false
  warning: false
  message: false
---

::: {.callout-note}
## Status and interpretation

**Archived implementation, not current evidence.** Its units, assumptions and
low-level execution predate the current contract. Do not use its results to
describe present HUB behaviour. Use `docs/sampling_variance_evaluation.qmd`.

This is an executable pilot study of **Monte Carlo sampling variance in the
current MIAMA-HUB implementation**. It is not yet a validation of structural
model uncertainty, health-model parameter uncertainty, or the
representativeness of the Leeds synthetic population.

The full design contains `r params$n_replicates * (length(strsplit(as.character(params$effect_percent), ",", fixed = TRUE)[[1]]) + as.integer(params$include_no_change)) * 2 * 2`
runs. Each replicate uses a different random seed. A master seed makes the set
of replicate seeds reproducible; it does not fix every assessment to the same
sample.
:::

# Research questions

This report evaluates four questions.

1. How much variation in total HALYs gained is introduced solely by the random
   selection of people and trips in the current counterfactual algorithm?
2. Does absolute or relative sampling variation change as the requested active
   travel increase grows from 10% to 100%?
3. Does variation differ between walking and cycling, whose reference user and
   trip pools differ substantially?
4. Does variation differ when an assessment is entered as additional weekly
   **users** rather than additional weekly **trips**?

The experiment changes one mode and one input unit at a time. For example, the
walking-user scenario increases walking users but does not independently impose
a walking-trip target. This isolates each counterfactual handler and the
downstream trip and physical-activity changes it generates.

# Experimental design

## Factorial scenarios

| Factor | Levels |
|---|---|
| Dataset | Packaged Leeds profile: 5,000 synthetic people |
| Mode | Walking; cycling |
| Active-travel input unit | Distinct weekly users; weighted weekly trips |
| Requested increase | No change, then +10%, +20%, ..., +100% relative to the reference value |
| Replicates | `r params$n_replicates` independently seeded runs per scenario |
| Primary outcome | Total HALYs gained over a 10-year POC assessment period |

For user scenarios, the target is rounded to a whole person because HUB samples
individual rows. For trip scenarios, the target remains a numeric weighted
weekly trip total. HUB converts the target into physical rows using the current
trip-count implementation; the realized weighted total may therefore differ
slightly from the requested target.

## Outcome and estimands

The primary run-level outcome is:

\[
\text{HALYs gained} = w_p \sum_{i,t=1}^{T} (HALY_{cf,i,t}-HALY_{ref,i,t}),
\]

where $w_p$ is the effective person weight for the Leeds sample and $T$ is the
configured assessment period. Cycle 0 is excluded because it is the initial
state rather than a modeled follow-up year.

For each mode × input-unit × effect-size scenario, the report calculates:

- mean and median HALYs gained;
- standard deviation (SD), interquartile range (IQR), and full range;
- a 95% empirical interval based on replicate quantiles;
- coefficient of variation, `100 * SD / abs(mean)`;
- signal-to-noise ratio, `abs(mean) / SD`;
- the probability that a replicate has a sign opposite to the scenario mean;
- requested and realized changes in users or trips; and
- MMET exposure change as an intermediate diagnostic.

With only ten replicates, quantiles, CVs, and sign probabilities are descriptive
diagnostics rather than precise uncertainty estimates.

# Simulation implementation

## Setup

```{r}
#| label: setup
#| include: false

project_candidates <- c(".", "..")
project_root <- project_candidates[
  vapply(project_candidates, function(path) {
    desc <- file.path(path, "DESCRIPTION")
    file.exists(desc) && identical(unname(read.dcf(desc)[1, "Package"]), "MIAMAHUB")
  }, logical(1))
][1]

if (is.na(project_root)) {
  stop("Render from the MIAMA-HUB package root or its docs directory.", call. = FALSE)
}
project_root <- normalizePath(project_root, winslash = "/", mustWork = TRUE)

# Evaluate the current checkout, not an older installed package.
if (!requireNamespace("devtools", quietly = TRUE)) {
  stop("The development report requires the devtools package.", call. = FALSE)
}
devtools::load_all(project_root, quiet = TRUE)

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
})

effect_percent <- as.numeric(strsplit(as.character(params$effect_percent), ",", fixed = TRUE)[[1]])
if (any(!is.finite(effect_percent)) || any(effect_percent <= 0)) {
  stop("effect_percent must contain positive comma-separated numbers.", call. = FALSE)
}
design_effect_percent <- if (isTRUE(params$include_no_change)) c(0, effect_percent) else effect_percent
sample_sizes <- as.integer(strsplit(as.character(params$sample_sizes), ",", fixed = TRUE)[[1]])
sample_size_effect_percent <- as.numeric(
  strsplit(as.character(params$sample_size_effect_percent), ",", fixed = TRUE)[[1]]
)
combined_effect_percent <- as.numeric(
  strsplit(as.character(params$combined_effect_percent), ",", fixed = TRUE)[[1]]
)
if (any(!is.finite(sample_sizes)) || any(sample_sizes < 2) ||
    any(!is.finite(sample_size_effect_percent)) || any(sample_size_effect_percent <= 0) ||
    any(!is.finite(combined_effect_percent)) || any(combined_effect_percent <= 0)) {
  stop("Sample-size parameters must contain valid positive comma-separated numbers.", call. = FALSE)
}
if (params$n_replicates < 2) {
  stop("n_replicates must be at least 2 to estimate variance.", call. = FALSE)
}

mode_specs <- list(
  walking = list(suffix = "walk", label = "Walking"),
  cycling = list(suffix = "bike", label = "Cycling")
)
units <- c("users", "trips")

design <- expand.grid(
  mode = names(mode_specs),
  at_data_unit = units,
  effect_percent = design_effect_percent,
  replicate = seq_len(params$n_replicates),
  stringsAsFactors = FALSE
) |>
  arrange(.data$at_data_unit, .data$mode, .data$effect_percent, .data$replicate)

set.seed(as.integer(params$master_seed))
design$seed <- sample.int(.Machine$integer.max, nrow(design), replace = FALSE)
stopifnot(!anyDuplicated(design$seed))
design$scenario_id <- sprintf(
  "%s_%s_p%03d_r%03d",
  design$mode,
  design$at_data_unit,
  as.integer(design$effect_percent),
  design$replicate
)

cache_dir <- file.path(
  project_root, "docs", "_sampling_variance_cache", params$cache_version
)
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)
```

## Load the common reference data once

All runs use the same reference population and health-enriched reference data.
Only the stochastic counterfactual sampling and resulting health lookup are
repeated. This both isolates the source of variation and avoids repeatedly
reading identical parquet files.

```{r}
#| label: reference-data

field <- function(value) {
  list(is_filled = TRUE, input_value = value)
}

profile <- MIAMAHUB:::build_mock_appraisal_inputs(
  overrides = list(
    ui_version = field("basic"),
    geo_level = field("lad"),
    geo_id = field(params$geography_id),
    modes = field(c("walking", "cycling")),
    at_data_unit = field("trips"),
    users_timeframe_walk = field("week"),
    users_timeframe_bike = field("week"),
    trips_timeframe_walk = field("week"),
    trips_timeframe_bike = field("week"),
    trips_denominator_walk = field("total"),
    trips_denominator_bike = field("total"),
    res_aggregation = field("timeline"),
    res_outcomes = field("halys")
  )
)

cfg <- MIAMAHUB::miama_default_config(dataset_size = params$dataset_size)
cfg$results$assessment_period_years <- as.integer(params$assessment_period_years)
hub <- MIAMAHUB::Hub$new(cfg = cfg)
profile <- hub$build_reference_profile_defaults(profile)
hub$build_reference_data_from_defaults()

reference_data <- hub$reference_data
reference_updates <- hub$reference_default_ui_values$ui_updates
base_values <- MIAMAHUB::receive_appraisal_inputs(profile)$appraisal_input_values
assessment_period <- MIAMAHUB::get_assessment_period(cfg)
person_weight <- cfg$population$person_weight

# Load only the POC cycle horizon once. These immutable tables are reused by
# every replicate; only counterfactual exposure and sampling vary by run.
census_ids <- unique(stats::na.omit(reference_data$ind$census_id))
hm_cycle_outcomes <- MIAMAHUB:::load_hm_cycle_outcomes_death_share(
  cfg,
  census_ids = census_ids,
  cycles = 0:assessment_period
)
reference_exposure <- MIAMAHUB:::.counterfactual_health_exposure(
  reference_data$ind,
  reference_data$ind
)
lookup_scope <- MIAMAHUB:::.counterfactual_health_lookup_scope(
  hm_cycle_outcomes,
  reference_exposure
)
hm_cycle_lookup <- MIAMAHUB:::load_hm_cycle_lookup_death_share(
  cfg,
  strata = lookup_scope$strata,
  cycles = lookup_scope$cycles
)

derive_reference_levels <- function(reference_data_run) {
  bind_rows(lapply(names(mode_specs), function(mode) {
  suffix <- mode_specs[[mode]]$suffix
  mode_spec <- MIAMAHUB:::.miama_tab2_mode_specs()[[mode]]
  active_trips <- mode_spec$trip_filter(reference_data_run$trips) &
    !is.na(reference_data_run$trips$nts_tripid)
  all_active <- MIAMAHUB:::.miama_tab2_mode_specs()$walking$trip_filter(reference_data_run$trips) |
    MIAMAHUB:::.miama_tab2_mode_specs()$cycling$trip_filter(reference_data_run$trips)
  broad_trip_candidates <- !all_active & !is.na(reference_data_run$trips$nts_tripid)
  if ("trip_utilitarian" %in% names(reference_data_run$trips)) {
    broad_trip_candidates <- broad_trip_candidates &
      MIAMAHUB:::.true_values(reference_data_run$trips$trip_utilitarian)
  }
  active_users <- MIAMAHUB:::.positive_col(reference_data_run$ind, mode_spec$ind_duration_col)
  data.frame(
    mode = mode,
    mode_label = mode_specs[[mode]]$label,
    reference_users = sum(active_users, na.rm = TRUE),
    reference_trips = sum(
      MIAMAHUB:::.trip_weights(reference_data_run$trips)[active_trips],
      na.rm = TRUE
    ),
    reference_active_trip_rows = sum(active_trips, na.rm = TRUE),
    eligible_new_users = sum(!active_users, na.rm = TRUE),
    broad_eligible_trip_rows = sum(broad_trip_candidates, na.rm = TRUE),
    stringsAsFactors = FALSE
  )
  }))
}

reference_levels <- derive_reference_levels(reference_data)

# Ensure the direct derivation used by controlled subsamples agrees with the
# production UI reference extract for the complete Leeds profile.
for (mode in names(mode_specs)) {
  suffix <- mode_specs[[mode]]$suffix
  row <- reference_levels[reference_levels$mode == mode, , drop = FALSE]
  stopifnot(isTRUE(all.equal(
    row$reference_users,
    as.numeric(reference_updates[[paste0("users_count_ref_", suffix)]])
  )))
  stopifnot(isTRUE(all.equal(
    row$reference_trips,
    as.numeric(reference_updates[[paste0("trips_count_ref_", suffix)]])
  )))
}

reference_levels
```

The effective person weight is `r round(person_weight, 3)`, and the configured
assessment period is `r assessment_period` years.

## One independently seeded assessment

The function below calls the production counterfactual and health-outcome
functions. It deliberately retains only compact run-level diagnostics and then
releases the large counterfactual object.

```{r}
#| label: simulation-function
#| echo: true

run_sampling_assessment <- function(mode,
                                    at_data_unit,
                                    effect_percent,
                                    replicate,
                                    seed,
                                    reference_data_run = reference_data,
                                    reference_levels_run = reference_levels,
                                    person_weight_run = person_weight,
                                    hm_cycle_outcomes_run = hm_cycle_outcomes,
                                    experiment = "main") {
  started <- proc.time()[["elapsed"]]
  spec <- mode_specs[[mode]]
  suffix <- spec$suffix
  ref_row <- reference_levels_run[reference_levels_run$mode == mode, , drop = FALSE]
  values <- base_values
  values$modes <- mode
  values$at_data_unit <- at_data_unit

  no_change <- identical(as.numeric(effect_percent), 0)
  if (no_change) {
    reference_target <- if (identical(at_data_unit, "users")) {
      ref_row$reference_users
    } else {
      ref_row$reference_trips
    }
    requested_target <- reference_target
  } else if (identical(at_data_unit, "users")) {
    reference_target <- ref_row$reference_users
    requested_target <- as.integer(round(reference_target * (1 + effect_percent / 100)))
    requested_target <- max(as.integer(reference_target + 1), requested_target)
    values[[paste0("users_count_cf_", suffix)]] <- requested_target
  } else {
    reference_target <- ref_row$reference_trips
    requested_target <- reference_target * (1 + effect_percent / 100)
    values[[paste0("trips_timeframe_", suffix)]] <- "week"
    values[[paste0("trips_denominator_", suffix)]] <- "total"
    values[[paste0("trips_count_cf_", suffix)]] <- requested_target
  }

  counterfactual_data <- MIAMAHUB:::init_counterfactual_data(reference_data_run)
  counterfactual_data <- MIAMAHUB:::apply_counterfactual_ui_values(
    counterfactual_data = counterfactual_data,
    appraisal_input_values = values,
    reference_data = reference_data_run,
    constants = MIAMAHUB:::miama_counterfactual_defaults(cfg),
    seed = as.integer(seed)
  )

  mode_spec <- MIAMAHUB:::.miama_tab2_mode_specs()[[mode]]
  if (identical(at_data_unit, "users")) {
    realized_target <- sum(
      MIAMAHUB:::.positive_col(counterfactual_data$ind, mode_spec$ind_duration_col),
      na.rm = TRUE
    )
  } else {
    active <- mode_spec$trip_filter(counterfactual_data$trips) &
      !is.na(counterfactual_data$trips$nts_tripid)
    realized_target <- sum(
      MIAMAHUB:::.trip_weights(counterfactual_data$trips)[active],
      na.rm = TRUE
    )
  }

  active_rows <- mode_spec$trip_filter(counterfactual_data$trips) &
    !is.na(counterfactual_data$trips$nts_tripid)
  realized_physical_trip_rows <- sum(active_rows, na.rm = TRUE)

  comparison <- counterfactual_data$counterfactual_report$comparison
  changed_individuals <- nrow(comparison$changed_ind_rows)
  changed_trip_rows <- nrow(comparison$changed_trip_rows)
  selection_keys <- c(
    if ("census_id" %in% names(comparison$changed_ind_rows)) {
      paste0("i:", comparison$changed_ind_rows$census_id)
    },
    if ("nts_tripid" %in% names(comparison$changed_trip_rows)) {
      paste0("t:", comparison$changed_trip_rows$nts_tripid)
    }
  )
  selection_keys <- sort(unique(selection_keys))
  selection_code <- if (length(selection_keys) == 0) {
    "none"
  } else {
    code_points <- utf8ToInt(paste(selection_keys, collapse = "|"))
    as.character(sum(code_points * (seq_along(code_points) %% 1000003)) %% 2147483647)
  }

  exposure <- counterfactual_data$counterfactual_report$mmet_exposure
  total_mmet_delta <- exposure$total_mmet_wkhr_delta[exposure$component == "total"]
  if (length(total_mmet_delta) == 0) total_mmet_delta <- NA_real_

  counterfactual_data <- MIAMAHUB:::apply_counterfactual_health_outcomes(
    counterfactual_data = counterfactual_data,
    reference_data = reference_data_run,
    cfg = cfg,
    scheme_effect_duration = "longterm",
    hm_cycle_outcomes = hm_cycle_outcomes_run,
    hm_cycle_lookup = hm_cycle_lookup
  )

  health <- counterfactual_data$health_outcomes
  keep <- is.finite(health$cycle) & health$cycle > 0 & health$cycle <= assessment_period
  halys_gained <- sum(health$d_haly[keep], na.rm = TRUE) * person_weight_run

  data.frame(
    mode = mode,
    mode_label = spec$label,
    at_data_unit = at_data_unit,
    unit_label = if (at_data_unit == "users") "Distinct weekly users" else "Weighted weekly trips",
    effect_percent = effect_percent,
    replicate = replicate,
    seed = seed,
    experiment = experiment,
    synthetic_sample_size = nrow(reference_data_run$ind),
    effective_person_weight = person_weight_run,
    reference_target = reference_target,
    requested_target = requested_target,
    realized_target = realized_target,
    requested_absolute_change = requested_target - reference_target,
    realized_absolute_change = realized_target - reference_target,
    realized_effect_percent = 100 * (realized_target / reference_target - 1),
    reference_active_trip_rows = ref_row$reference_active_trip_rows,
    realized_active_trip_rows = realized_physical_trip_rows,
    eligible_new_users = ref_row$eligible_new_users,
    broad_eligible_trip_rows = ref_row$broad_eligible_trip_rows,
    changed_individuals = changed_individuals,
    changed_trip_rows = changed_trip_rows,
    selection_code = selection_code,
    total_mmet_wkhr_delta = total_mmet_delta,
    halys_gained = halys_gained,
    halys_per_mmet_wkhr = ifelse(total_mmet_delta != 0, halys_gained / total_mmet_delta, NA_real_),
    elapsed_seconds = proc.time()[["elapsed"]] - started,
    stringsAsFactors = FALSE
  )
}
```

## Execute or resume the factorial simulation

Each run is saved separately. Re-rendering resumes from completed runs when
`reuse_cached_runs` is true. Increment `cache_version` after changing model code
or the simulation definition; otherwise old and new results could be mixed.

```{r}
#| label: execute-simulation

run_one_design_row <- function(row) {
  cache_path <- file.path(cache_dir, paste0(row$scenario_id, ".rds"))
  if (isTRUE(params$reuse_cached_runs) && file.exists(cache_path)) {
    return(readRDS(cache_path))
  }
  if (!isTRUE(params$run_simulation)) {
    return(NULL)
  }

  result <- run_sampling_assessment(
    mode = row$mode,
    at_data_unit = row$at_data_unit,
    effect_percent = row$effect_percent,
    replicate = row$replicate,
    seed = row$seed
  )
  temporary_path <- paste0(cache_path, ".tmp")
  saveRDS(result, temporary_path)
  if (!file.rename(temporary_path, cache_path)) {
    stop("Could not finalize simulation cache: ", cache_path, call. = FALSE)
  }
  result
}

simulation_rows <- lapply(seq_len(nrow(design)), function(i) {
  result <- run_one_design_row(design[i, , drop = FALSE])
  if (i %% 10 == 0 || i == nrow(design)) {
    message("Sampling simulation: ", i, " / ", nrow(design), " runs checked")
  }
  result
})
simulation_rows <- Filter(Negate(is.null), simulation_rows)

if (length(simulation_rows) == 0) {
  stop(
    "No simulation results are available. Set run_simulation: true or provide cached runs.",
    call. = FALSE
  )
}
simulation_results <- bind_rows(simulation_rows)

if (nrow(simulation_results) != nrow(design)) {
  warning(
    "The report is incomplete: ", nrow(simulation_results), " of ", nrow(design),
    " requested runs are available."
  )
}

# Structural checks: varying seeds must be unique, while no-change runs must
# remain exactly unchanged regardless of seed or selected handler context.
if (anyDuplicated(simulation_results$seed)) {
  stop("Simulation output contains duplicated random seeds.", call. = FALSE)
}
no_change_results <- simulation_results[simulation_results$effect_percent == 0, , drop = FALSE]
if (nrow(no_change_results) > 0) {
  invariant_failed <- abs(no_change_results$halys_gained) > 1e-10 |
    abs(no_change_results$total_mmet_wkhr_delta) > 1e-10 |
    no_change_results$changed_individuals != 0 |
    no_change_results$changed_trip_rows != 0
  if (any(invariant_failed)) {
    stop("No-change invariant failed for one or more independently seeded runs.", call. = FALSE)
  }
}
```

# Results

## Target fidelity

This diagnostic distinguishes variance in the requested sampling mechanism from
failure to attain the requested user or trip total.

```{r}
#| label: target-fidelity

target_fidelity <- simulation_results |>
  group_by(.data$mode_label, .data$unit_label, .data$effect_percent) |>
  summarise(
    requested_change = mean(.data$requested_absolute_change),
    mean_realized_change = mean(.data$realized_absolute_change),
    mean_realized_effect_percent = mean(.data$realized_effect_percent),
    max_absolute_target_error = max(abs(.data$realized_target - .data$requested_target)),
    mean_changed_individuals = mean(.data$changed_individuals),
    mean_changed_trip_rows = mean(.data$changed_trip_rows),
    .groups = "drop"
  )

target_fidelity
```

## Randomization and no-change audits

Different seeds are necessary but not sufficient evidence that different rows
were sampled. `selection_code` is a compact checksum of changed person and trip
identifiers. It is used only as an audit diagnostic, not as a random seed.

```{r}
#| label: randomization-audit

selection_audit <- simulation_results |>
  filter(.data$effect_percent > 0) |>
  group_by(.data$mode_label, .data$unit_label, .data$effect_percent) |>
  summarise(
    replicates = n(),
    distinct_seeds = n_distinct(.data$seed),
    distinct_changed_row_sets = n_distinct(.data$selection_code),
    min_changed_individuals = min(.data$changed_individuals),
    max_changed_individuals = max(.data$changed_individuals),
    min_changed_trip_rows = min(.data$changed_trip_rows),
    max_changed_trip_rows = max(.data$changed_trip_rows),
    .groups = "drop"
  )

selection_audit
```

```{r}
#| label: no-change-audit

if (nrow(no_change_results) > 0) {
  no_change_results |>
    summarise(
      runs = n(),
      maximum_absolute_haly = max(abs(.data$halys_gained)),
      maximum_absolute_mmet_delta = max(abs(.data$total_mmet_wkhr_delta)),
      changed_individuals = sum(.data$changed_individuals),
      changed_trip_rows = sum(.data$changed_trip_rows)
    )
}
```

All no-change quantities must be exactly zero within numerical tolerance. The
report stops during execution rather than merely warning if this invariant is
violated.

## Replicate-level HALY impacts

Each point is one independently seeded assessment. The line is the scenario
mean; the shaded interval spans the minimum and maximum among the available
replicates. A wider vertical spread at a given effect size indicates greater
Monte Carlo sensitivity to which people or trips were sampled.

```{r}
#| label: impact-points
#| fig-width: 10
#| fig-height: 7

impact_summary <- simulation_results |>
  group_by(.data$mode_label, .data$unit_label, .data$effect_percent) |>
  summarise(
    mean_haly = mean(.data$halys_gained),
    min_haly = min(.data$halys_gained),
    max_haly = max(.data$halys_gained),
    .groups = "drop"
  )

ggplot(simulation_results, aes(x = .data$effect_percent, y = .data$halys_gained)) +
  geom_ribbon(
    data = impact_summary,
    aes(x = .data$effect_percent, ymin = .data$min_haly, ymax = .data$max_haly),
    inherit.aes = FALSE,
    fill = "#B8D8CF", alpha = 0.45
  ) +
  geom_line(data = impact_summary, aes(y = .data$mean_haly), color = "#006B57", linewidth = 0.8) +
  geom_point(
    aes(group = factor(.data$replicate)),
    position = position_jitter(width = 1.2, height = 0),
    alpha = 0.65, size = 1.7, color = "#314A59"
  ) +
  facet_grid(.data$unit_label ~ .data$mode_label, scales = "free_y") +
  scale_x_continuous(breaks = effect_percent) +
  labs(
    title = "Total HALYs gained across independently sampled assessments",
    subtitle = "Points are replicates; line is the mean; ribbon is the replicate range",
    x = "Requested increase from reference (%)",
    y = paste0("HALYs gained over ", assessment_period, " years")
  ) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

## Scenario-level sampling spread

```{r}
#| label: variance-summary

variance_summary <- simulation_results |>
  group_by(.data$mode, .data$mode_label, .data$at_data_unit, .data$unit_label, .data$effect_percent) |>
  summarise(
    n = n(),
    mean_haly = mean(.data$halys_gained),
    median_haly = median(.data$halys_gained),
    sd_haly = sd(.data$halys_gained),
    iqr_haly = IQR(.data$halys_gained),
    min_haly = min(.data$halys_gained),
    max_haly = max(.data$halys_gained),
    q025_haly = unname(quantile(.data$halys_gained, 0.025)),
    q975_haly = unname(quantile(.data$halys_gained, 0.975)),
    negative_rate = mean(.data$halys_gained < 0),
    mean_realized_effect_percent = mean(.data$realized_effect_percent),
    mean_mmet_delta = mean(.data$total_mmet_wkhr_delta),
    .groups = "drop"
  ) |>
  mutate(
    cv_percent = ifelse(abs(.data$mean_haly) > 0, 100 * .data$sd_haly / abs(.data$mean_haly), NA_real_),
    signal_to_noise = ifelse(.data$sd_haly > 0, abs(.data$mean_haly) / .data$sd_haly, Inf),
    opposite_sign_rate = ifelse(
      .data$mean_haly > 0,
      .data$negative_rate,
      ifelse(.data$mean_haly < 0, 1 - .data$negative_rate, NA_real_)
    )
  )

variance_summary
```

## Centered views that expose within-scenario variance

The dose-response relationship can visually overwhelm replicate variation.
The next plot removes each scenario mean and expresses every run as a percentage
deviation from that mean. It should be read alongside absolute SD because a
large percentage deviation can still represent a small number of HALYs.

```{r}
#| label: centered-variation-plot
#| fig-width: 11
#| fig-height: 7

centered_results <- simulation_results |>
  filter(.data$effect_percent > 0) |>
  group_by(.data$mode_label, .data$unit_label, .data$effect_percent) |>
  mutate(
    scenario_mean_haly = mean(.data$halys_gained),
    haly_deviation = .data$halys_gained - .data$scenario_mean_haly,
    relative_deviation_percent = 100 * .data$haly_deviation / abs(.data$scenario_mean_haly)
  ) |>
  ungroup()

ggplot(
  centered_results,
  aes(x = factor(.data$effect_percent), y = .data$relative_deviation_percent)
) +
  geom_hline(yintercept = 0, color = "grey50", linewidth = 0.35) +
  geom_boxplot(width = 0.55, outlier.shape = NA, fill = "#DCECE7", color = "#006B57") +
  geom_jitter(width = 0.14, height = 0, alpha = 0.65, size = 1.5, color = "#314A59") +
  facet_grid(.data$unit_label ~ .data$mode_label, scales = "free_y") +
  labs(
    title = "Replicate deviation from each scenario's mean HALY impact",
    subtitle = "Centering removes the dominant increase in mean impact across effect sizes",
    x = "Requested increase from reference (%)",
    y = "Deviation from scenario mean (%)"
  ) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

```{r}
#| label: absolute-centered-variation-plot
#| fig-width: 11
#| fig-height: 7

ggplot(
  centered_results,
  aes(x = factor(.data$effect_percent), y = .data$haly_deviation)
) +
  geom_hline(yintercept = 0, color = "grey50", linewidth = 0.35) +
  geom_boxplot(width = 0.55, outlier.shape = NA, fill = "#E6EAF0", color = "#314A59") +
  geom_jitter(width = 0.14, height = 0, alpha = 0.65, size = 1.5, color = "#8C2F5C") +
  facet_grid(.data$unit_label ~ .data$mode_label, scales = "free_y") +
  labs(
    title = "Absolute replicate deviation from scenario mean",
    x = "Requested increase from reference (%)",
    y = "HALY deviation from scenario mean"
  ) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

## Uncertainty in the estimated standard deviation

With ten replicates, the SD itself is uncertain. This non-parametric bootstrap
resamples the available replicate outcomes within each scenario. It does not
replace additional model runs, but it shows whether an apparently precise SD is
supported by the number of replicates.

```{r}
#| label: sd-bootstrap

bootstrap_sd_interval <- function(values, draws, seed) {
  values <- as.numeric(values)
  if (length(values) < 2 || draws < 1) return(c(low = NA_real_, high = NA_real_))
  set.seed(seed)
  estimates <- replicate(draws, stats::sd(sample(values, replace = TRUE)))
  stats::quantile(estimates, c(0.025, 0.975), names = FALSE, na.rm = TRUE) |>
    setNames(c("low", "high"))
}

scenario_groups <- split(
  simulation_results[simulation_results$effect_percent > 0, , drop = FALSE],
  interaction(
    simulation_results$mode[simulation_results$effect_percent > 0],
    simulation_results$at_data_unit[simulation_results$effect_percent > 0],
    simulation_results$effect_percent[simulation_results$effect_percent > 0],
    drop = TRUE
  )
)

sd_uncertainty <- bind_rows(lapply(seq_along(scenario_groups), function(index) {
  group <- scenario_groups[[index]]
  interval <- bootstrap_sd_interval(
    group$halys_gained,
    draws = as.integer(params$sd_bootstrap_draws),
    seed = as.integer(params$master_seed) + index
  )
  data.frame(
    mode_label = group$mode_label[1],
    unit_label = group$unit_label[1],
    effect_percent = group$effect_percent[1],
    sd_haly = sd(group$halys_gained),
    sd_low = interval[["low"]],
    sd_high = interval[["high"]]
  )
}))

ggplot(sd_uncertainty, aes(x = .data$effect_percent, y = .data$sd_haly)) +
  geom_ribbon(aes(ymin = .data$sd_low, ymax = .data$sd_high), fill = "#E7CAD8", alpha = 0.55) +
  geom_line(color = "#8C2F5C", linewidth = 0.8) +
  geom_point(color = "#8C2F5C", size = 1.8) +
  facet_grid(.data$unit_label ~ .data$mode_label, scales = "free_y") +
  labs(
    title = "Estimated HALY SD and bootstrap uncertainty",
    subtitle = paste0(params$sd_bootstrap_draws, " bootstrap resamples within each scenario"),
    x = "Requested increase from reference (%)",
    y = "SD of total HALYs gained"
  ) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

```{r}
#| label: absolute-spread-plot
#| fig-width: 10
#| fig-height: 6

ggplot(variance_summary, aes(x = .data$effect_percent, y = .data$sd_haly)) +
  geom_line(color = "#006B57", linewidth = 0.8) +
  geom_point(color = "#006B57", size = 2) +
  facet_grid(.data$unit_label ~ .data$mode_label, scales = "free_y") +
  scale_x_continuous(breaks = effect_percent) +
  labs(
    title = "Absolute Monte Carlo spread in total HALYs",
    subtitle = "Standard deviation across independently seeded assessments",
    x = "Requested increase from reference (%)",
    y = "SD of total HALYs gained"
  ) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

## Sampling spread relative to effect size

Absolute SD may rise with a larger intervention even while uncertainty becomes
small relative to the mean effect. CV and signal-to-noise therefore answer a
different question than SD.

```{r}
#| label: relative-spread-plot
#| fig-width: 10
#| fig-height: 6

ggplot(variance_summary, aes(x = .data$effect_percent, y = .data$cv_percent)) +
  geom_hline(yintercept = c(5, 10, 20), linetype = "dotted", color = "grey65") +
  geom_line(color = "#8C2F5C", linewidth = 0.8) +
  geom_point(color = "#8C2F5C", size = 2) +
  facet_grid(.data$unit_label ~ .data$mode_label, scales = "free_y") +
  scale_x_continuous(breaks = effect_percent) +
  labs(
    title = "Sampling spread relative to mean HALY impact",
    subtitle = "CV = 100 × SD / absolute mean; high values indicate a weak signal relative to Monte Carlo variation",
    x = "Requested increase from reference (%)",
    y = "Coefficient of variation (%)"
  ) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

## Relationship to the number of additional sampled units

This plot places the variation against the absolute requested increase rather
than its percentage. Walking and cycling can then be compared in terms of the
actual number of added users or weighted trips represented by a scenario.

```{r}
#| label: size-spread-plot
#| fig-width: 10
#| fig-height: 5

size_summary <- simulation_results |>
  group_by(.data$mode_label, .data$unit_label, .data$effect_percent) |>
  summarise(
    additional_units = mean(.data$requested_absolute_change),
    mean_haly = mean(.data$halys_gained),
    sd_haly = sd(.data$halys_gained),
    .groups = "drop"
  ) |>
  mutate(
    cv_percent = ifelse(abs(.data$mean_haly) > 0, 100 * .data$sd_haly / abs(.data$mean_haly), NA_real_)
  )

ggplot(size_summary, aes(x = .data$additional_units, y = .data$cv_percent, color = .data$mode_label)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_wrap(vars(.data$unit_label), scales = "free_x") +
  labs(
    title = "Relative sampling spread by absolute intervention size",
    x = "Requested additional users or weighted weekly trips",
    y = "Coefficient of variation (%)",
    color = "Mode"
  ) +
  scale_color_manual(values = c("Walking" = "#006B57", "Cycling" = "#8C2F5C")) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(), legend.position = "top")
```

## Baseline donor pools and candidate saturation

The same percentage increase can require very different numbers of sampled
rows. These denominators identify scenarios approaching the available donor
pool, where fallback rules or deterministic inclusion of most candidates can
reduce apparent random variation.

```{r}
#| label: donor-pool-summary

reference_levels |>
  select(
    .data$mode_label, .data$reference_users, .data$eligible_new_users,
    .data$reference_active_trip_rows, .data$reference_trips,
    .data$broad_eligible_trip_rows
  )
```

```{r}
#| label: candidate-saturation-data

saturation_data <- simulation_results |>
  filter(.data$effect_percent > 0) |>
  mutate(
    donor_pool = ifelse(
      .data$at_data_unit == "users",
      .data$eligible_new_users,
      .data$broad_eligible_trip_rows
    ),
    changed_rows = ifelse(
      .data$at_data_unit == "users",
      .data$changed_individuals,
      .data$changed_trip_rows
    ),
    candidate_pool_used_percent = 100 * .data$changed_rows / .data$donor_pool
  )
```

The candidate-pool relationship is summarized below using the already centered
HALY results. The broad trip pool is an upper bound: final eligibility also
depends on plausible distance and other active constraints.

```{r}
#| label: candidate-saturation-final-plot
#| fig-width: 10
#| fig-height: 5

saturation_plot_data <- saturation_data |>
  select(
    .data$mode_label, .data$unit_label, .data$effect_percent,
    .data$replicate, .data$candidate_pool_used_percent
  ) |>
  left_join(
    centered_results |>
      select(
        .data$mode_label, .data$unit_label, .data$effect_percent,
        .data$replicate, .data$relative_deviation_percent
      ),
    by = c("mode_label", "unit_label", "effect_percent", "replicate")
  )

ggplot(
  saturation_plot_data,
  aes(x = .data$candidate_pool_used_percent, y = abs(.data$relative_deviation_percent))
) +
  geom_point(alpha = 0.55, color = "#314A59") +
  geom_smooth(method = "loess", se = FALSE, color = "#006B57", linewidth = 0.8) +
  facet_grid(.data$unit_label ~ .data$mode_label, scales = "free") +
  labs(
    title = "Relative HALY deviation versus share of broad candidate pool changed",
    x = "Broad candidate pool used (%)",
    y = "Absolute deviation from scenario mean (%)"
  ) +
  theme_minimal(base_size = 10)
```

## MMET-exposure versus health-response variation

Random sampling can affect HALYs through at least two pathways: the amount of
MMET exposure generated and the health trajectories of the people receiving
that exposure. Within-scenario regression on total MMET change provides a first
decomposition. Residual variation includes variation in who receives exposure,
nonlinearity, and any information not represented by the aggregate MMET total;
it is not independent health-model randomness, because the health lookup itself
is deterministic conditional on the sampled counterfactual.

```{r}
#| label: mmet-haly-decomposition

decomposition_groups <- split(
  simulation_results[simulation_results$effect_percent > 0, , drop = FALSE],
  interaction(
    simulation_results$mode[simulation_results$effect_percent > 0],
    simulation_results$at_data_unit[simulation_results$effect_percent > 0],
    simulation_results$effect_percent[simulation_results$effect_percent > 0],
    drop = TRUE
  )
)

health_response_decomposition <- bind_rows(lapply(decomposition_groups, function(group) {
  estimable <- nrow(group) >= 3 &&
    length(unique(group$total_mmet_wkhr_delta[is.finite(group$total_mmet_wkhr_delta)])) >= 2
  fit <- if (estimable) stats::lm(halys_gained ~ total_mmet_wkhr_delta, data = group) else NULL
  data.frame(
    mode_label = group$mode_label[1],
    unit_label = group$unit_label[1],
    effect_percent = group$effect_percent[1],
    mmet_haly_correlation = if (estimable) stats::cor(group$total_mmet_wkhr_delta, group$halys_gained) else NA_real_,
    mmet_r_squared = if (estimable) summary(fit)$r.squared else NA_real_,
    residual_haly_sd = if (estimable) stats::sd(stats::residuals(fit)) else NA_real_,
    total_haly_sd = stats::sd(group$halys_gained),
    stringsAsFactors = FALSE
  )
}))

health_response_decomposition
```

```{r}
#| label: mmet-haly-plot
#| fig-width: 10
#| fig-height: 7

ggplot(
  simulation_results |> filter(.data$effect_percent > 0),
  aes(
    x = .data$total_mmet_wkhr_delta,
    y = .data$halys_gained,
    color = .data$effect_percent
  )
) +
  geom_point(alpha = 0.65, size = 1.5) +
  facet_grid(.data$unit_label ~ .data$mode_label, scales = "free") +
  scale_color_viridis_c(option = "C", end = 0.85) +
  labs(
    title = "HALYs gained versus total weekly MMET change",
    subtitle = "Color denotes the requested percentage increase",
    x = "Total MMET-hours/week change across synthetic people",
    y = paste0("HALYs gained over ", assessment_period, " years"),
    color = "Increase (%)"
  ) +
  theme_minimal(base_size = 10)
```

```{r}
#| label: halys-per-mmet-plot
#| fig-width: 10
#| fig-height: 6

ggplot(
  simulation_results |> filter(.data$effect_percent > 0),
  aes(x = factor(.data$effect_percent), y = .data$halys_per_mmet_wkhr)
) +
  geom_boxplot(outlier.shape = NA, fill = "#DCECE7", color = "#006B57") +
  geom_jitter(width = 0.14, alpha = 0.55, size = 1.3, color = "#314A59") +
  facet_grid(.data$unit_label ~ .data$mode_label, scales = "free_y") +
  labs(
    title = "Health response per unit of aggregate MMET change",
    x = "Requested increase from reference (%)",
    y = "HALYs gained per MMET-hour/week"
  ) +
  theme_minimal(base_size = 10)
```

## Weighted-trip conversion diagnostics

For trip scenarios, weighted target error and the number of physical rows
changed are separate quantities. The table and plot below test whether larger
weighted-target errors coincide with larger departures from the scenario mean.

```{r}
#| label: weighted-trip-diagnostics

weighted_trip_diagnostics <- simulation_results |>
  filter(.data$at_data_unit == "trips", .data$effect_percent > 0) |>
  mutate(
    weighted_target_error = .data$realized_target - .data$requested_target,
    weighted_target_error_percent = 100 * .data$weighted_target_error / .data$requested_target
  ) |>
  left_join(
    centered_results |>
      select(
        .data$mode_label, .data$unit_label, .data$effect_percent,
        .data$replicate, .data$haly_deviation, .data$relative_deviation_percent
      ),
    by = c("mode_label", "unit_label", "effect_percent", "replicate")
  )

weighted_trip_diagnostics |>
  group_by(.data$mode_label, .data$effect_percent) |>
  summarise(
    mean_weighted_target_error_percent = mean(.data$weighted_target_error_percent),
    max_absolute_weighted_target_error_percent = max(abs(.data$weighted_target_error_percent)),
    mean_changed_trip_rows = mean(.data$changed_trip_rows),
    .groups = "drop"
  )
```

```{r}
#| label: weighted-trip-error-plot
#| fig-width: 9
#| fig-height: 4.5

ggplot(
  weighted_trip_diagnostics,
  aes(x = .data$weighted_target_error_percent, y = .data$relative_deviation_percent)
) +
  geom_hline(yintercept = 0, color = "grey70", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "grey70", linewidth = 0.3) +
  geom_point(aes(color = .data$effect_percent), alpha = 0.65) +
  facet_wrap(vars(.data$mode_label), scales = "free") +
  scale_color_viridis_c(option = "C", end = 0.85) +
  labs(
    title = "Weighted-trip target error versus HALY deviation",
    x = "Realized minus requested weighted trips (%)",
    y = "HALY deviation from scenario mean (%)",
    color = "Increase (%)"
  ) +
  theme_minimal(base_size = 10)
```

## Expected stabilization from averaging independent seeds

If independent seeded runs are averaged, their Monte Carlo standard error is
expected to decrease approximately with $1/\sqrt{k}$. This is an analytical
projection from the observed replicate SD, not a substitute for validating
independence or running the model repeatedly.

```{r}
#| label: seed-averaging-plot
#| fig-width: 10
#| fig-height: 6

average_counts <- c(1, 2, 5, 10, 20, 50)
seed_averaging <- bind_rows(lapply(average_counts, function(k) {
  transform(
    variance_summary[variance_summary$effect_percent > 0, , drop = FALSE],
    averaged_runs = k
  )
})) |>
  mutate(expected_cv_of_mean = .data$cv_percent / sqrt(.data$averaged_runs))

ggplot(
  seed_averaging,
  aes(x = .data$averaged_runs, y = .data$expected_cv_of_mean, group = .data$effect_percent)
) +
  geom_line(alpha = 0.45, color = "#314A59") +
  geom_point(alpha = 0.6, size = 1.2, color = "#006B57") +
  facet_grid(.data$unit_label ~ .data$mode_label, scales = "free_y") +
  scale_x_continuous(breaks = average_counts) +
  labs(
    title = "Projected relative Monte Carlo error after averaging independent runs",
    subtitle = "Each line represents one requested effect size",
    x = "Number of independently seeded runs averaged",
    y = "Expected CV of averaged impact (%)"
  ) +
  theme_minimal(base_size = 10)
```

# Optional controlled sample-size experiment

The main design varies intervention size but keeps the 5,000-person synthetic
sample fixed. The optional experiment below uses nested subsamples of the same
Leeds profile to distinguish intervention effect size from the number of
synthetic people available to the sampler.

Each smaller sample is rescaled to represent the same Leeds population. This
tests Monte Carlo precision under reduced synthetic resolution; it does **not**
simulate a genuinely smaller town. One fixed nested reference subsample is used
per size so the first comparison isolates counterfactual sampling conditional
on that reference. Repeating the reference-subsample draw is a further required
refinement.

This experiment is disabled by default because its configured design adds
`r length(sample_sizes) * length(sample_size_effect_percent) * 2 * 2 * params$sample_size_replicates`
health-model runs. Enable it with:

```sh
quarto render docs/sampling_variance_evaluation.qmd --to html \
  -P run_sample_size_experiment:true
```

```{r}
#| label: sample-size-experiment

sample_size_results <- data.frame()

if (isTRUE(params$run_sample_size_experiment)) {
  if (max(sample_sizes) > nrow(reference_data$ind)) {
    stop("A requested sample size exceeds the available Leeds profile.", call. = FALSE)
  }
  if (params$sample_size_replicates < 2) {
    stop("sample_size_replicates must be at least 2.", call. = FALSE)
  }

  set.seed(as.integer(params$master_seed) + 50000L)
  nested_id_order <- sample(unique(reference_data$ind$census_id), replace = FALSE)
  sample_bundles <- lapply(sample_sizes, function(sample_n) {
    selected_ids <- nested_id_order[seq_len(sample_n)]
    reference_subset <- reference_data
    reference_subset$ind <- reference_data$ind[
      reference_data$ind$census_id %in% selected_ids, , drop = FALSE
    ]
    reference_subset$trips <- reference_data$trips[
      reference_data$trips$census_id %in% selected_ids, , drop = FALSE
    ]
    health_subset <- hm_cycle_outcomes[
      hm_cycle_outcomes$census_id %in% selected_ids, , drop = FALSE
    ]
    list(
      sample_size = sample_n,
      reference_data = reference_subset,
      reference_levels = derive_reference_levels(reference_subset),
      hm_cycle_outcomes = health_subset,
      # Preserve the represented Leeds population while reducing synthetic
      # resolution. This scaling does not affect CV but does affect absolute SD.
      person_weight = person_weight * nrow(reference_data$ind) / sample_n
    )
  })
  names(sample_bundles) <- as.character(sample_sizes)

  sample_design <- expand.grid(
    synthetic_sample_size = sample_sizes,
    mode = names(mode_specs),
    at_data_unit = units,
    effect_percent = sample_size_effect_percent,
    replicate = seq_len(params$sample_size_replicates),
    stringsAsFactors = FALSE
  ) |>
    arrange(
      .data$synthetic_sample_size, .data$at_data_unit, .data$mode,
      .data$effect_percent, .data$replicate
    )

  set.seed(as.integer(params$master_seed) + 60000L)
  sample_design$seed <- sample.int(
    .Machine$integer.max,
    nrow(sample_design),
    replace = FALSE
  )
  sample_design$scenario_id <- sprintf(
    "sample_n%05d_%s_%s_p%03d_r%03d",
    sample_design$synthetic_sample_size,
    sample_design$mode,
    sample_design$at_data_unit,
    as.integer(sample_design$effect_percent),
    sample_design$replicate
  )

  sample_rows <- lapply(seq_len(nrow(sample_design)), function(i) {
    row <- sample_design[i, , drop = FALSE]
    cache_path <- file.path(cache_dir, paste0(row$scenario_id, ".rds"))
    if (isTRUE(params$reuse_cached_runs) && file.exists(cache_path)) {
      return(readRDS(cache_path))
    }
    if (!isTRUE(params$run_simulation)) return(NULL)

    bundle <- sample_bundles[[as.character(row$synthetic_sample_size)]]
    result <- run_sampling_assessment(
      mode = row$mode,
      at_data_unit = row$at_data_unit,
      effect_percent = row$effect_percent,
      replicate = row$replicate,
      seed = row$seed,
      reference_data_run = bundle$reference_data,
      reference_levels_run = bundle$reference_levels,
      person_weight_run = bundle$person_weight,
      hm_cycle_outcomes_run = bundle$hm_cycle_outcomes,
      experiment = "controlled_sample_size"
    )
    temporary_path <- paste0(cache_path, ".tmp")
    saveRDS(result, temporary_path)
    if (!file.rename(temporary_path, cache_path)) {
      stop("Could not finalize sample-size cache: ", cache_path, call. = FALSE)
    }
    result
  })

  sample_rows <- Filter(Negate(is.null), sample_rows)
  if (length(sample_rows) > 0) sample_size_results <- bind_rows(sample_rows)
}

if (nrow(sample_size_results) == 0) {
  cat("Controlled sample-size experiment not run in this render.\n")
}
```

```{r}
#| label: sample-size-results
#| fig-width: 11
#| fig-height: 7

if (nrow(sample_size_results) > 0) {
  sample_size_summary <- sample_size_results |>
    group_by(
      .data$synthetic_sample_size, .data$mode_label, .data$unit_label,
      .data$effect_percent
    ) |>
    summarise(
      mean_haly = mean(.data$halys_gained),
      sd_haly = sd(.data$halys_gained),
      mean_changed_individuals = mean(.data$changed_individuals),
      mean_changed_trip_rows = mean(.data$changed_trip_rows),
      .groups = "drop"
    ) |>
    mutate(
      cv_percent = 100 * .data$sd_haly / abs(.data$mean_haly)
    )

  print(
    ggplot(
      sample_size_summary,
      aes(
        x = .data$synthetic_sample_size,
        y = .data$cv_percent,
        color = factor(.data$effect_percent),
        group = factor(.data$effect_percent)
      )
    ) +
      geom_line(linewidth = 0.75) +
      geom_point(size = 1.8) +
      facet_grid(.data$unit_label ~ .data$mode_label, scales = "free_y") +
      scale_x_log10(breaks = sample_sizes) +
      labs(
        title = "Relative HALY sampling variation by synthetic sample size",
        subtitle = "Nested Leeds subsamples rescaled to the same represented population",
        x = "Synthetic people available (log scale)",
        y = "Coefficient of variation (%)",
        color = "Increase (%)"
      ) +
      theme_minimal(base_size = 10)
  )
}
```

# Optional combined-mode and ordering experiment

Walking and cycling compete for some of the same non-active trip candidates.
The production sampler locks a shifted trip so it cannot subsequently switch
again, but the composition of the remaining candidate pool can still depend on
which mode is processed first. This paired experiment applies the same
percentage increase to both modes, runs both processing orders with the same
seed, and compares the combined impact with separate-mode results.

It is disabled by default. Its standard configuration adds
`r length(combined_effect_percent) * 2 * 2 * params$combined_replicates` runs.

```{r}
#| label: combined-mode-experiment

run_combined_assessment <- function(at_data_unit,
                                    effect_percent,
                                    replicate,
                                    seed,
                                    mode_order) {
  started <- proc.time()[["elapsed"]]
  values <- base_values
  values$modes <- mode_order
  values$at_data_unit <- at_data_unit

  targets <- list()
  for (mode in mode_order) {
    spec <- mode_specs[[mode]]
    ref_row <- reference_levels[reference_levels$mode == mode, , drop = FALSE]
    if (identical(at_data_unit, "users")) {
      reference_target <- ref_row$reference_users
      target <- max(
        as.integer(reference_target + 1),
        as.integer(round(reference_target * (1 + effect_percent / 100)))
      )
      values[[paste0("users_count_cf_", spec$suffix)]] <- target
    } else {
      reference_target <- ref_row$reference_trips
      target <- reference_target * (1 + effect_percent / 100)
      values[[paste0("trips_timeframe_", spec$suffix)]] <- "week"
      values[[paste0("trips_denominator_", spec$suffix)]] <- "total"
      values[[paste0("trips_count_cf_", spec$suffix)]] <- target
    }
    targets[[mode]] <- c(reference = reference_target, requested = target)
  }

  counterfactual_data <- MIAMAHUB:::init_counterfactual_data(reference_data)
  counterfactual_data <- MIAMAHUB:::apply_counterfactual_ui_values(
    counterfactual_data = counterfactual_data,
    appraisal_input_values = values,
    reference_data = reference_data,
    constants = MIAMAHUB:::miama_counterfactual_defaults(cfg),
    seed = as.integer(seed)
  )

  realized <- vapply(names(mode_specs), function(mode) {
    mode_spec <- MIAMAHUB:::.miama_tab2_mode_specs()[[mode]]
    if (identical(at_data_unit, "users")) {
      sum(
        MIAMAHUB:::.positive_col(counterfactual_data$ind, mode_spec$ind_duration_col),
        na.rm = TRUE
      )
    } else {
      active <- mode_spec$trip_filter(counterfactual_data$trips) &
        !is.na(counterfactual_data$trips$nts_tripid)
      sum(MIAMAHUB:::.trip_weights(counterfactual_data$trips)[active], na.rm = TRUE)
    }
  }, numeric(1))

  exposure <- counterfactual_data$counterfactual_report$mmet_exposure
  total_mmet_delta <- exposure$total_mmet_wkhr_delta[exposure$component == "total"]
  comparison <- counterfactual_data$counterfactual_report$comparison
  counterfactual_data <- MIAMAHUB:::apply_counterfactual_health_outcomes(
    counterfactual_data = counterfactual_data,
    reference_data = reference_data,
    cfg = cfg,
    scheme_effect_duration = "longterm",
    hm_cycle_outcomes = hm_cycle_outcomes,
    hm_cycle_lookup = hm_cycle_lookup
  )
  health <- counterfactual_data$health_outcomes
  keep <- is.finite(health$cycle) & health$cycle > 0 & health$cycle <= assessment_period

  data.frame(
    at_data_unit = at_data_unit,
    unit_label = if (at_data_unit == "users") "Distinct weekly users" else "Weighted weekly trips",
    effect_percent = effect_percent,
    replicate = replicate,
    seed = seed,
    mode_order = paste(mode_order, collapse = " then "),
    walking_reference = targets$walking[["reference"]],
    walking_requested = targets$walking[["requested"]],
    walking_realized = realized[["walking"]],
    cycling_reference = targets$cycling[["reference"]],
    cycling_requested = targets$cycling[["requested"]],
    cycling_realized = realized[["cycling"]],
    changed_individuals = nrow(comparison$changed_ind_rows),
    changed_trip_rows = nrow(comparison$changed_trip_rows),
    total_mmet_wkhr_delta = total_mmet_delta,
    halys_gained = sum(health$d_haly[keep], na.rm = TRUE) * person_weight,
    elapsed_seconds = proc.time()[["elapsed"]] - started,
    stringsAsFactors = FALSE
  )
}

combined_results <- data.frame()
if (isTRUE(params$run_combined_mode_experiment)) {
  if (params$combined_replicates < 2) {
    stop("combined_replicates must be at least 2.", call. = FALSE)
  }
  paired_design <- expand.grid(
    at_data_unit = units,
    effect_percent = combined_effect_percent,
    replicate = seq_len(params$combined_replicates),
    stringsAsFactors = FALSE
  )
  set.seed(as.integer(params$master_seed) + 70000L)
  paired_design$seed <- sample.int(.Machine$integer.max, nrow(paired_design), replace = FALSE)
  combined_design <- bind_rows(lapply(seq_len(nrow(paired_design)), function(i) {
    row <- paired_design[i, , drop = FALSE]
    bind_rows(lapply(c("walking_first", "cycling_first"), function(order_id) {
      transform(row, order_id = order_id)
    }))
  }))
  combined_design$scenario_id <- sprintf(
    "combined_%s_%s_p%03d_r%03d",
    combined_design$order_id,
    combined_design$at_data_unit,
    as.integer(combined_design$effect_percent),
    combined_design$replicate
  )

  combined_rows <- lapply(seq_len(nrow(combined_design)), function(i) {
    row <- combined_design[i, , drop = FALSE]
    cache_path <- file.path(cache_dir, paste0(row$scenario_id, ".rds"))
    if (isTRUE(params$reuse_cached_runs) && file.exists(cache_path)) {
      return(readRDS(cache_path))
    }
    if (!isTRUE(params$run_simulation)) return(NULL)
    mode_order <- if (row$order_id == "walking_first") {
      c("walking", "cycling")
    } else {
      c("cycling", "walking")
    }
    result <- run_combined_assessment(
      at_data_unit = row$at_data_unit,
      effect_percent = row$effect_percent,
      replicate = row$replicate,
      seed = row$seed,
      mode_order = mode_order
    )
    temporary_path <- paste0(cache_path, ".tmp")
    saveRDS(result, temporary_path)
    if (!file.rename(temporary_path, cache_path)) {
      stop("Could not finalize combined-mode cache: ", cache_path, call. = FALSE)
    }
    result
  })
  combined_rows <- Filter(Negate(is.null), combined_rows)
  if (length(combined_rows) > 0) combined_results <- bind_rows(combined_rows)
}

if (nrow(combined_results) == 0) {
  cat("Combined-mode experiment not run in this render.\n")
}
```

```{r}
#| label: combined-mode-results
#| fig-width: 10
#| fig-height: 5

if (nrow(combined_results) > 0) {
  order_comparison <- combined_results |>
    select(
      .data$unit_label, .data$effect_percent, .data$replicate,
      .data$mode_order, .data$halys_gained
    ) |>
    reshape(
      idvar = c("unit_label", "effect_percent", "replicate"),
      timevar = "mode_order",
      direction = "wide"
    )
  order_columns <- grep("^halys_gained\\.", names(order_comparison), value = TRUE)
  order_comparison$order_difference_haly <-
    order_comparison[[order_columns[1]]] - order_comparison[[order_columns[2]]]

  print(
    ggplot(
      order_comparison,
      aes(x = factor(.data$effect_percent), y = .data$order_difference_haly)
    ) +
      geom_hline(yintercept = 0, color = "grey50", linewidth = 0.35) +
      geom_boxplot(outlier.shape = NA, fill = "#DCECE7", color = "#006B57") +
      geom_jitter(width = 0.14, alpha = 0.65, color = "#314A59") +
      facet_wrap(vars(.data$unit_label), scales = "free_y") +
      labs(
        title = "Paired effect of walking-versus-cycling processing order",
        subtitle = "Each pair uses the same random seed and combined targets",
        x = "Requested increase in both modes (%)",
        y = "Difference in combined HALYs between processing orders"
      ) +
      theme_minimal(base_size = 10)
  )

  separate_means <- simulation_results |>
    filter(.data$effect_percent %in% combined_effect_percent) |>
    group_by(.data$unit_label, .data$effect_percent, .data$mode) |>
    summarise(mean_haly = mean(.data$halys_gained), .groups = "drop") |>
    group_by(.data$unit_label, .data$effect_percent) |>
    summarise(sum_separate_mode_haly = sum(.data$mean_haly), .groups = "drop")
  combined_interaction <- combined_results |>
    group_by(.data$unit_label, .data$effect_percent, .data$mode_order) |>
    summarise(mean_combined_haly = mean(.data$halys_gained), .groups = "drop") |>
    left_join(separate_means, by = c("unit_label", "effect_percent")) |>
    mutate(
      interaction_haly = .data$mean_combined_haly - .data$sum_separate_mode_haly,
      interaction_percent = 100 * .data$interaction_haly / abs(.data$sum_separate_mode_haly)
    )

  combined_interaction
}
```

# Interpretation framework

The following diagnostics should be considered together:

- **Target fidelity:** Large requested-versus-realized discrepancies indicate a
  target-conversion or candidate-pool issue, not merely health-impact variance.
- **Absolute SD:** Answers how many HALYs the result changes when the random
  sample changes.
- **CV:** Answers whether that spread is large compared with the estimated
  impact. It becomes unstable when the mean is near zero.
- **Signal-to-noise:** Values below 1 mean replicate SD exceeds the mean effect;
  values above 5 or 10 indicate increasingly stable rank/order conclusions.
- **Sign reversals:** Any sign changes for an intervention expected to improve
  health require inspection of sampled exposure and health trajectories.
- **MMET spread:** Helps distinguish transport-sampling variation from
  nonlinear variation introduced by the health lookup.

No universal acceptable CV threshold is imposed here. A threshold should be
chosen with reference to the decision being supported and the smallest policy-
relevant HALY difference, not solely statistical convention.

# Limitations of this pilot

1. The Leeds profile has 5,000 synthetic rows but represents the full Leeds
   population through an effective person weight. The weight scales both means
   and SDs; it does not create additional independent sampled people.
2. Ten replicates provide a first view but estimate tails and SD imprecisely.
   Bootstrap intervals now expose this imprecision, while the replicate count
   remains configurable for targeted 50–100-run follow-up scenarios.
3. Relative intervention size is not the same as statistical sample size.
   Cycling may have a much smaller reference pool than walking even at the same
   percentage increase.
4. The optional controlled sample-size experiment varies synthetic resolution
   but uses one fixed nested reference draw per size. Multiple independent
   reference draws are still needed to include reference-sample composition
   uncertainty and it does not represent genuinely smaller towns.
5. Only HALYs are assessed. Disease-specific outcomes may have higher variance,
   especially for uncommon outcomes or narrow demographic strata.
6. The health-model trajectories and HALY parameters are treated as fixed.
7. User and trip scenarios invoke different counterfactual mechanisms. Their
   variance should not be interpreted as a pure comparison of two equivalent
   measurement units.

# Recommended decision criteria

This pilot can identify scenarios that are clearly unstable, but a minimum
supported assessment size should be set only after the refinements below. A
candidate operational rule should combine:

- a minimum absolute number of changed synthetic people or trips;
- a maximum Monte Carlo CV for the primary outcome;
- a minimum signal-to-noise ratio;
- no material sign-reversal rate; and
- acceptable requested-versus-realized target error.

If a scenario fails the rule, the application could increase the number of
replicates, report a simulation interval, use a larger underlying synthetic
sample, or warn that the result is below the supported assessment scale.

# Refinement task list

- [x] Make replicate count configurable for targeted 50–100-run scenarios and
      quantify uncertainty in the estimated SD with a bootstrap interval.
- [x] Add a no-change scenario across seeds; it must return exactly zero and
      acts as an implementation invariant.
- [x] Add an optional nested-subsample design with 250, 500, 1,000, 2,500, and
      5,000 people to separate intervention effect size from synthetic sample
      size; next add independent reference draws at every size.
- [x] Stratify by baseline active-mode pool size and number of eligible donor
      people/trips, not only requested percentage.
- [x] Add an optional paired combined walking-and-cycling experiment to assess
      ordering and cross-mode candidate competition.
- [ ] Compare unconstrained sampling with each advanced age, sex, PA, distance,
      purpose, and diversion constraint separately.
- [x] Decompose variance into aggregate MMET-exposure variation and residual
      health-response variation; next extend beyond aggregate MMET totals.
- [ ] Repeat for mortality and selected common/rare disease outcomes.
- [ ] Test short-term versus long-term scheme-effect duration.
- [x] Evaluate whether weighted-trip target error covaries with HALY variation
      beyond selection of physical trip rows.
- [ ] Define policy-relevant absolute HALY tolerances before choosing a CV or
      signal-to-noise threshold.
- [x] Project the expected reduction in Monte Carlo error from averaging
      independent seeds; production adoption still requires a threshold and
      run-level audit design.

# Reproducibility record

```{r}
#| label: reproducibility

data.frame(
  item = c(
    "Dataset", "Geography", "Synthetic people", "Reference trip rows",
    "Person weight", "Assessment period", "Requested runs", "Completed runs",
    "Completed sample-size runs", "Completed combined-mode runs",
    "Master seed", "Cache version"
  ),
  value = c(
    params$dataset_size,
    params$geography_id,
    nrow(reference_data$ind),
    nrow(reference_data$trips),
    person_weight,
    assessment_period,
    nrow(design),
    nrow(simulation_results),
    nrow(sample_size_results),
    nrow(combined_results),
    params$master_seed,
    params$cache_version
  ),
  stringsAsFactors = FALSE
)
```

```{r}
#| label: session-info
#| echo: true

sessionInfo()
```
