---
date: "2026-09-10"
title: "Users versus trips: a matched-snapshot experiment"
format:
  html:
    toc: true
    embed-resources: true
params:
  data_dir: "_route_benchmark"
execute:
  echo: false
  warning: false
---

# What this comparison can show

We start with one fixed sample of people and their linked trips. For each mode,
we construct a counterfactual by switching existing short car trips. We then
ask HUB to reconstruct that scenario from either its user counts or its trip
counts. Neither reconstruction is supplied with the other route's count target.

Differences **between repeated seeds** measure sampling variation conditional
on this fixed population. Persistent differences **between routes** can reflect
different completion/allocation assumptions, missing information, or defects;
they are not automatically bugs. Equal total MMET does not imply equal health
benefits if different people receive the activity.

```{r}
library(dplyr)
library(ggplot2)
runs <- read.csv(file.path(params$data_dir, "runs.csv"))
runs$warnings[is.na(runs$warnings)] <- ""
bench <- read.csv(file.path(params$data_dir, "benchmarks.csv"))
unavailable <- read.csv(file.path(params$data_dir, "unavailable.csv"))
manifest <- readRDS(file.path(params$data_dir, "manifest.rds"))
ok <- filter(runs, status == "ok")
knitr::kable(data.frame(Population = manifest$population_size,
  Seeds = length(manifest$seeds), Successful = nrow(ok), Failed = sum(runs$status != "ok")))
```

# Design and boundaries

- All modes: walking, cycling, e-biking, and PT access walking.
- Two benchmarks: new users (previously no AT in any mode), and extra trips
  among current users of the target mode. Up to five eligible people receive up
  to two switched trips each. Donor journeys are positive-distance, at most
  5 km, and utilitarian. The benchmark does not add rows or induce trips.
- Fixed population identities, no geographic population multiplier, 40 model
  years and immediate full effect for every route.
- **Defaults:** normal completion assumptions, except shared zero induction and
  car-only diversion. **Benchmark-informed:** trip frequency and distance are
  derived from the actual switched trips; new-user preference is 100% or 0%
  according to scenario. These settings still do not encode recipient identities.
- **Basic:** direct final-results API. **Advanced:** Tab 2 population
  staging followed by acceptance of the generated Tab 3 tables and Tab 4
  staging. Both now use `Hub$build_results()` with canonical persisted
  assumptions. This is not a browser test. Earlier rendered results must be
  regenerated after this workflow change before interpreting them as current.
- Seeds are shared across routes for convenience; identical seeds do not
  guarantee identical people are sampled. Source selection uses a separate,
  fixed seed. This pilot does not measure uncertainty in the source sample or HM.

An existing-users benchmark can have identical REF/CF user counts but additional
trips. The users-only route cannot convey that increase. A zero result on that
route is an expected information limitation, not evidence of sampling failure.
Native e-bike current-user scenarios are unavailable if the sample has no e-bikers.

The benchmark's new users have no prior AT in any mode. Explicit user-count
requests, however, can recruit people who already use another AT mode; they
override the new-user percentage. Consequently, even benchmark-informed
aggregate assumptions do not guarantee the same recipient pool or dose per
person. This is one of the allocation distinctions the experiment is meant to
expose, rather than hide by forcing identical recipients into both routes.

## Benchmarks and unavailable cases

```{r}
knitr::kable(select(bench, mode, scenario, ref_users, users, ref_trips, trips,
                    trip_minutes, mmet_delta, halys), digits = 3)
if (nrow(unavailable)) knitr::kable(unavailable)
```

# Health results

## First things to inspect

```{r}
#| results: asis
keys <- c("mode", "scenario", "route", "assumptions", "seed")
paired <- merge(filter(ok, workflow == "basic"), filter(ok, workflow == "advanced"), by = keys)
if (nrow(paired)) cat("- Largest matched basic/advanced HALY difference: **",
  format(max(abs(paired$halys.x - paired$halys.y)), digits = 4), "**. Both paths include final results reconstruction.\n", sep = "")
cat("- For extra trips among existing users, the users-only inputs are unchanged. Their zero-change result is expected.\n")
cat("- Compare the two assumption settings before attributing a route difference to randomness. Even matching trip counts need not match person-level activity or baseline risk.\n")
cat("- Walking and PT trips remain distinct. PT walkers qualify as existing walking recipients; their eligibility alone does not add them to separate-walking user counts.\n")
```

Black horizontal lines show benchmark HALYs. Points are individual seeds;
colours distinguish assumption settings. The panels deliberately have separate
scales. This refresh uses `r length(manifest$seeds)` draws per combination;
it is a diagnostic comparison, not a reliable confidence interval.

::: {.callout-warning title="PT correctness finding in this refresh"}
The two-seed current-code refresh matches basic/advanced results exactly for
walking, cycling and e-biking, but not PT. For the PT trips route, Tab 2 targets
of 85 REF / 94 CF trips become generated Tab 4 defaults of zero; the final
advanced result then contains zero CF PT trips. The largest paired HALY
difference is 1.429874. Treat PT results below as diagnostic evidence of this
defect, not sampling variation. A targeted replay confirms that staged data
contains 85/94 PT access-walking trips, while the materialized REF/CF views
contain zero. Shared walking-component columns are cleared during scope
materialization. Preserve PT access activity there and add a real-data
regression before rerunning.
:::

```{r}
#| fig-width: 12
#| fig-height: 13
plot_runs <- ok |> mutate(path = factor(paste(workflow, route, sep = " / "),
  levels = c("basic / users", "basic / trips", "advanced / users", "advanced / trips")))
ggplot(plot_runs, aes(path, halys, colour = assumptions)) +
  geom_hline(data = bench, aes(yintercept = halys), inherit.aes = FALSE) +
  geom_point(position = position_jitter(width = 0.12, height = 0, seed = 1), alpha = 0.6) +
  facet_wrap(vars(mode, scenario), scales = "free_y", ncol = 2,
    labeller = labeller(mode = c(walking = "Walking", cycling = "Cycling", ebiking = "E-biking", pt = "PT access walking"),
      scenario = c(new_users = "New users", existing_users = "Extra trips among existing users"))) +
  scale_colour_manual(values = c(defaults = "#167568", benchmark_informed = "#B7406B"),
    breaks = c("defaults", "benchmark_informed"), labels = c("Default assumptions", "Benchmark-informed")) +
  labs(x = NULL, y = "HALYs gained over 40 years", colour = "Assumptions") +
  theme_minimal() + theme(legend.position = "bottom", axis.text.x = element_text(angle = 25, hjust = 1))
```

## Route means and variation

```{r}
summary <- ok |> group_by(mode, scenario, workflow, route, assumptions) |>
  summarise(n = n(), mean_halys = mean(halys), sd_halys = sd(halys),
    mean_difference_from_benchmark = mean(haly_difference),
    distinct_exposures = n_distinct(exposure_hash), .groups = "drop")
knitr::kable(summary, digits = 4)
```

# Where differences arise

Inspect counts first, then exposure, then health. `person_mmet_mae` is the mean
absolute difference from the benchmark's person-level MMET changes, including
unchanged people. Trip minutes and individual-record minutes are shown separately:
HUB can assign exposure through different components, so those measures must not
be added together as if independent. `mmet_delta` is the reconciled exposure.

```{r}
audit <- ok |> group_by(mode, scenario, workflow, route, assumptions) |>
  summarise(users = mean(users), trips = mean(trips),
    trip_minutes = mean(trip_minutes), individual_minutes = mean(individual_minutes),
    mmet_delta = mean(mmet_delta), changed_people = mean(changed_people),
    person_mmet_mae = mean(person_mmet_mae),
    max_copied_people = max(donor_copies), max_extra_trip_rows = max(extra_trip_rows),
    .groups = "drop")
knitr::kable(audit, digits = 3)
```

## Errors and fallbacks

Failed runs are retained, not excluded without explanation. Added trip rows or
copies mean the reconstruction used a fallback outside the benchmark's simple
shift-only mechanism; interpret that difference separately from seed variance.

```{r}
knitr::kable(runs |> filter(status != "ok" | nzchar(warnings)) |>
  select(any_of(c("mode", "scenario", "workflow", "route", "assumptions", "seed", "status", "error", "warnings"))))
```

# Reproduce and inspect

From the HUB root, with the project's R dependencies available:

```bash
Rscript inst/workflows/dev_route_benchmark.R 10 500
quarto render docs/current/validation/route_benchmark.qmd
```

Outputs in `docs/current/validation/_route_benchmark/` include CSV summaries, exact benchmark
snapshots and per-run profiles/exposure records, and a manifest with code checksums,
source/health/lookup hashes, configuration and seeds. Runs are recomputed rather
than reusing the older variance report's caches. This directory is git-ignored;
the rendered report is the shareable overview. The code checksum captures local
uncommitted changes as well as the recorded Git commit.

```{r}
knitr::kable(data.frame(Item = c("Run time", "HUB commit", "Source fingerprint", "Zero-change HALYs"),
  Value = c(as.character(manifest$timestamp), manifest$commit, manifest$source_hash,
            as.character(manifest$zero_change_health["halys"]))))
```
