---
date: "2026-09-10"
title: "Sampling variation in the current HUB pipeline"
format:
  html:
    toc: true
    embed-resources: true
    code-fold: true
  pdf:
    toc: true
params:
  data_dir: "_sampling_variance_current"
  population_data_dir: "_sampling_variance_current"
execute:
  echo: false
  warning: false
---

# Scope

This replaces the archived low-level pilot. It uses canonical persisted
`assump_*` fields, Tab 2/3/4 staging and real `Hub$build_results()` aggregation.
One record is one appraisal person/trip under the current contract. **Whether
the source trip weights need to be applied during source preparation remains
under review.** Sampling variance cannot establish representativeness or correct
systematic differences caused by ignoring survey weights.

The experiment holds supplied counts, assumptions, source and HM data fixed.
The 40-year scheme has immediate, permanent effect. The normal configured
new-user and induced-trip settings apply. REF trips are measured from one
observed population; the trip targets below remain unchanged across replicate
seeds. The population-size study requests 20% more trips. The cycling-volume
study instead holds all 5,000 source people fixed and requests specified
absolute increases of 100, 500 and 1,000 weekly cycling trips.

- **Whole appraisal:** repeat REF selection and CF allocation.
- **Fixed REF:** preserve one REF person/trip snapshot and repeat CF allocation.

These are complementary experiments, not an additive variance decomposition.
Fixed REF conditions on one particular cohort, so its mean may differ from the
whole-appraisal mean. Its SD need not be smaller in a short pilot. Population-size
comparisons also use different observed trip totals and completion assumptions;
they are realistic scenario comparisons, not an isolated sample-size experiment.
Neither tests browser save/discard state. User/trip route sensitivity is examined
separately in [the matched-route report](../../archive/2026-09-11/route_benchmark.html).

```{r}
library(dplyr)
library(ggplot2)
runs <- read.csv(file.path(params$data_dir, "runs.csv"))
summary <- read.csv(file.path(params$data_dir, "summary.csv")) |> mutate(variance = sd^2)
diagnostics <- read.csv(file.path(params$data_dir, "diagnostics.csv"))
manifest <- readRDS(file.path(params$data_dir, "manifest.rds"))
assumptions <- readRDS(file.path(params$data_dir, "assumptions.rds"))
source("../../../inst/workflows/dev_sampling_variance.R")
trip_inputs <- sv_trip_inputs(runs, params$data_dir)
scenario_keys <- sv_scenario_keys(trip_inputs)
mode_labels <- c(walking = "Walking", cycling = "Cycling", ebiking = "E-biking", pt = "PT access walking")
trip_context <- trip_inputs |>
  mutate(label = paste0(mode_labels[mode], ": ", ref_weekly_trips, " -> ",
    cf_weekly_trips, " (", ifelse(additional_weekly_trips >= 0, "+", ""),
    additional_weekly_trips, "/week)")) |>
  group_by(across(all_of(scenario_keys))) |>
  summarise(trip_context = paste(label, collapse = "; "),
    scenario_label = paste0(first(population), " people\n", paste(label, collapse = "\n")),
    .groups = "drop")
summary_with_trips <- left_join(summary, select(trip_context, -scenario_label), by = scenario_keys)
write.csv(trip_inputs, file.path(params$data_dir, "scenario_trip_inputs.csv"), row.names = FALSE)
write.csv(summary_with_trips, file.path(params$data_dir, "summary_with_trip_context.csv"), row.names = FALSE)
knitr::kable(data.frame(Replicates = manifest$n_replicates,
  Successful = n_distinct(runs$id[runs$status == "ok"]),
  Failed = n_distinct(runs$id[runs$status != "ok"])))
```

```{r}
#| results: asis
if (identical(manifest$design, "cycling_trip_increments")) {
  cat("**This report is the cycling-trip-volume study:** all ", manifest$sizes,
    " source people are assessed in each scenario; only the requested cycling-trip increase changes. ",
    "These are not population-expanded estimates for the entire LAD. ",
    "The earlier small-population study is retained separately.\n\n", sep = "")
  h <- filter(runs, status == "ok", outcome == "halys")
  paired <- merge(filter(h, experiment == "whole_appraisal"),
    filter(h, experiment == "fixed_ref"), by = c("additional_cycling_trips", "seed"))
  if (nrow(paired)) cat("Largest paired HALY difference between calculation paths: **",
    format(max(abs(paired$benefit.x - paired$benefit.y)), digits = 4),
    "**. The paths are an implementation check; do not pool them as independent draws.\n\n", sep = "")
}
```

# Trip volumes behind the scenarios

Population size alone does not describe the intervention being tested. These
are the **supplied weekly trip targets**, recovered from every successful run's
saved profile and checked for consistency across seeds and both experiments.
They are not estimates recalculated from the population size, nor a new audit
of realized trip rows. Combined scenarios list each mode separately.

```{r}
knitr::kable(select(trip_inputs, population, modes, mode, ref_weekly_trips,
  cf_weekly_trips, additional_weekly_trips), col.names = c("People", "Scenario", "Mode",
  "REF trips/week", "CF trips/week", "Additional trips/week"))
```

```{r}
#| results: asis
cycling_inputs <- filter(trip_inputs, modes == "cycling")
if (nrow(cycling_inputs)) {
  x <- cycling_inputs[which.min(cycling_inputs$additional_weekly_trips), ]
  cat("In the smallest cycling-trip increase (", x$population, " people), **",
    x$ref_weekly_trips, " REF trips become ", x$cf_weekly_trips, " CF trips: ",
    x$additional_weekly_trips, " additional weekly trips**.\n\n", sep = "")
}
```

Walking and cycling scenarios with the same population do not necessarily
describe equally sized interventions. Compare their additional trip volumes
before interpreting differences in health-result variability.

Trip targets are fixed, but recipients and sampled activity doses can differ.
The reported relative SD is **100 x SD(health benefit) / abs(mean health benefit)**,
not SD of the trip count or SD divided by trips. A large percentage can describe
a small absolute fluctuation around a small health benefit. Dividing all health
results by the same fixed number of additional trips would not change this
relative SD; it would only change the units of the mean and absolute SD.

# Health-result variation

Read the mean together with the SD: a small absolute difference can be a large
percentage when the expected benefit is close to zero. Start with HALYs below,
then inspect which recipients and activity doses differ in the saved diagnostics.
Do not interpret the spread between the two experiment means as random error.

Benefits use unrounded `all_modes` health-cube values over cycles 1-40:
CF minus REF for HALYs; REF minus CF for deaths and disease events. Do not sum
different outcome units. Relative SD is suppressed near zero mean benefit.
With fewer than 100 repetitions, tail ranges and sign frequencies are pilot
diagnostics. Empirical sampling ranges are not confidence intervals for true
health effects, and exclude uncertainty in assumptions and HM parameters.
Each plot panel has its own y-axis scale. Compare numerical SDs and relative
SDs, not the visual height of the point clouds across different panels.

**Colour key:** green/teal means **Fixed REF** (repeat CF allocation from one
unchanged REF); pink/red means **Whole appraisal** (repeat staging and CF
allocation). These colours are **not REF versus CF**. Each dot is one run's
health benefit over 40 years: **absolute HALYs gained** or **absolute deaths
prevented**, not a rate per 100,000 and not a percentage. Fractional deaths
represent expected counts. The horizontal categories are the two experiments;
small sideways jitter only keeps dots visible and has no numerical meaning.

```{r, fig.width=if(knitr::is_latex_output()) 6.5 else 12, fig.height=if(knitr::is_latex_output()) 8 else 18}
plot_data <- runs |> filter(status == "ok", outcome %in% c("halys", "mortality")) |>
  left_join(trip_context, by = scenario_keys) |>
  mutate(scenario_label = factor(scenario_label, levels = .env$trip_context$scenario_label))
variance_plot <- function(data, columns) {
  ggplot(data, aes(experiment, benefit, colour = experiment)) +
  geom_point(position = position_jitter(width = .12, height = 0, seed = 1), alpha = .5) +
  facet_wrap(vars(scenario_label, outcome), scales = "free_y", ncol = columns,
    labeller = labeller(outcome = c(halys = "HALYs gained\n40 years", mortality = "Deaths prevented\n40 years"))) +
  scale_colour_manual(name = "Resampling experiment",
    values = c(fixed_ref = "#287D70", whole_appraisal = "#C44976"),
    labels = c(fixed_ref = "Fixed REF", whole_appraisal = "Whole appraisal")) +
  scale_x_discrete(labels = c(fixed_ref = "Fixed REF", whole_appraisal = "Whole appraisal")) +
  labs(x = NULL, y = "HALYs gained / deaths prevented") +
  theme_minimal(base_size = 11) + theme(legend.position = "bottom", legend.title = element_blank())
}
if (knitr::is_latex_output()) {
  for (size in sort(unique(plot_data$population))) {
    print(variance_plot(filter(plot_data, population == size), 2))
  }
} else {
  print(variance_plot(plot_data, 3))
}
```

```{r}
knitr::kable(summary_with_trips |> filter(outcome %in% c("halys", "mortality")) |>
  select(population, trip_context, experiment, outcome, n, mean, sd, relative_sd_percent),
  col.names = c("People", "Weekly trips: REF -> CF (change)", "Experiment", "Outcome",
    "Draws", "Mean benefit", "SD of benefit", "Relative SD (%)"), digits = 4)
```

::: {.callout-note}
These are fresh diagnostic runs, not the proposed final 100-replicate study.
The main comparison above concerns HALYs and mortality; all disease-specific
summaries remain in the generated `summary.csv` rather than being added together.
`summary_with_trip_context.csv` includes the same full statistics with mode-specific
trip context; `scenario_trip_inputs.csv` supplies the separate numeric trip columns.
:::

# Aggregated variability plots

The following plots use **whole-appraisal results only**, not a pooling of the
two calculation paths. Each point summarizes ten seeds in the current pilots.
Lines connect the tested scenarios; they are not fitted predictions or confidence
bands. We plot **SD** (in the outcome's original units) and **relative SD** (%)
rather than squared-unit variance; the exported summary also contains
`variance = sd^2` for analyses that require variance itself.

## Variability versus additional cycling trips

This comparison holds the assessed population at 5,000 people. Its horizontal
axis is **additional cycling trips per week**, not population size or number of
Monte Carlo repetitions. Relative SD has the same percentage scale across both
outcomes. In the second plot, bars show **mean plus/minus one SD**, not a
confidence interval. The y-axis scale is fixed across trip increments within
each outcome, but HALYs and deaths retain separate outcome scales.

```{r, fig.width=6.5, fig.height=3.8}
trip_scale <- filter(summary, experiment == "whole_appraisal",
  outcome %in% c("halys", "mortality"))
if ("additional_cycling_trips" %in% names(trip_scale)) {
  trip_scale <- mutate(trip_scale, outcome_label = factor(recode(outcome,
    halys = "HALYs gained", mortality = "Deaths prevented"),
    levels = c("HALYs gained", "Deaths prevented")))
  print(ggplot(trip_scale, aes(additional_cycling_trips, relative_sd_percent)) +
    geom_line(colour = "#34495E") + geom_point(size = 2.5, colour = "#34495E") +
    facet_wrap(vars(outcome_label), nrow = 1) +
    scale_x_continuous(breaks = sort(unique(trip_scale$additional_cycling_trips))) +
    expand_limits(y = 0) +
    labs(x = "Additional cycling trips per week", y = "Relative SD of benefit (%)") +
    theme_minimal(base_size = 11) + theme(panel.spacing.x = grid::unit(1.2, "lines")))
} else {
  cat("This batch varies population size, not cycling-trip increments. See the separate cycling-volume report for this comparison.")
}
```

```{r, fig.width=6.5, fig.height=3.8}
if ("additional_cycling_trips" %in% names(trip_scale)) {
  print(ggplot(trip_scale, aes(additional_cycling_trips, mean)) +
    geom_line(colour = "#34495E") + geom_point(size = 2.5, colour = "#34495E") +
    geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 35) +
    facet_wrap(vars(outcome_label), nrow = 1, scales = "free_y") +
    scale_x_continuous(breaks = sort(unique(trip_scale$additional_cycling_trips))) +
    expand_limits(y = 0) +
    labs(x = "Additional cycling trips per week", y = "40-year benefit: mean +/- SD") +
    theme_minimal(base_size = 11) + theme(panel.spacing.x = grid::unit(1.2, "lines")))
}
```

## Variability versus assessed population

This uses the **earlier 100/500/1,500-person study**, not the new fixed-5,000
design. Colour denotes the active modes in this plot, as labelled in its legend.
The horizontal axis is assessed people, not the number of repetitions. All
panels share the same relative-SD percentage scale.

**This is not a pure sample-size effect:** trip increments and completion
assumptions also differ between those scenarios. For example, cycling adds
5/14/41 weekly trips, whereas walking adds 74/512/1,470. Labels below show the
trip increases; do not interpret the lines as controlling for intervention size.

```{r, fig.width=6.5, fig.height=4.5}
population_file <- file.path(params$population_data_dir, "summary.csv")
if (file.exists(population_file)) {
  pop_summary <- read.csv(population_file) |>
    filter(experiment == "whole_appraisal", outcome %in% c("halys", "mortality")) |>
    mutate(outcome_label = factor(recode(outcome, halys = "HALYs gained", mortality = "Deaths prevented"),
      levels = c("HALYs gained", "Deaths prevented")))
  print(ggplot(pop_summary, aes(population, relative_sd_percent, colour = modes)) +
    geom_line() + geom_point(size = 2) +
    facet_wrap(vars(outcome_label), nrow = 1) +
    scale_x_continuous(breaks = sort(unique(pop_summary$population))) +
    scale_colour_manual(name = NULL,
      values = c(cycling = "#3274A1", walking = "#8064A2", "walking+cycling" = "#C77920"),
      labels = c(cycling = "Cycling", walking = "Walking", "walking+cycling" = "Both modes")) +
    expand_limits(y = 0) +
    labs(x = "Assessed people", y = "Relative SD of benefit (%)") +
    theme_minimal(base_size = 11) + theme(legend.position = "bottom", panel.spacing.x = grid::unit(1.2, "lines")))
  pop_runs <- read.csv(file.path(params$population_data_dir, "runs.csv"))
  pop_inputs <- sv_trip_inputs(pop_runs, params$population_data_dir)
  knitr::kable(select(pop_inputs, population, modes, mode, additional_weekly_trips),
    col.names = c("People", "Scenario", "Mode", "Additional trips/week"))
} else {
  cat("Earlier population-size results are not available in the configured directory; no replacement data have been inferred.")
}
```

# Diagnostics and failures

The diagnostics CSV contains changed-person counts, recipient age and baseline
activity, MMET changes, donor copies and extra trip rows. Run-level RDS files
retain profiles, person-level exposure, and the sampling report containing
current/new users, shifted/induced trips and donor fallbacks. Failure rows are
retained; successful-run summaries can be misleading if failures are systematic.

```{r}
knitr::kable(runs |> distinct(id, status, warnings) |>
  filter(status != "ok" | (!is.na(warnings) & nzchar(warnings))))
knitr::kable(data.frame(Generated = as.character(manifest$timestamp), Commit = manifest$commit))
```

# Reproduction

**Rendering the saved results does not rerun the simulation.** From `docs`:

```sh
quarto render sampling_variance_evaluation.qmd --to html
quarto render sampling_variance_evaluation.qmd --to pdf
```

Only to generate a new simulation batch, from HUB root:

```r
pkgload::load_all(".")
source("inst/workflows/dev_sampling_variance.R")
run_sampling_variance(n_replicates = 100)
```

For the larger cycling-trip study, hold the full source population fixed:

```r
run_sampling_variance(n_replicates = 10,
  additional_cycling_trips = c(100, 500, 1000),
  output_dir = "docs/current/validation/_sampling_variance_cycling_trips")
```

Render this batch by passing `-P data_dir:_sampling_variance_cycling_trips`
and a separate output name, such as `--output cycling_trip_scale_variance.pdf`.
The increases are total additional cycling trips, not shifted-only targets:
the unchanged induced-trip assumption still applies. With all source people
included, the whole-appraisal arm cannot measure geographic person-subset
variation; differences from fixed REF can still involve trip allocation or
staging. This is an intervention-volume comparison, not a larger source sample.

Render from `docs/current/validation`: `quarto render sampling_variance_evaluation.qmd`.
Runs are fresh, never reused from the old pilot cache. The manifest saves
configuration, seeds, source/code/data fingerprints and the R session.
The original implementation is retained under `docs/archive/`, not used here.

# Appendix: assumptions and provenance

Defaults are resolved once per scenario with `prepare_assumption_profile()`.
Every final profile and `get_appraisal_assumptions()` output is saved. Effective
values and provenance must have the same signature across repetitions, or the
run is recorded as an error. Card applicability may change after accepting
generated counts without the underlying values changing.

```{r}
rows <- bind_rows(lapply(names(assumptions), function(scenario) {
  bind_rows(lapply(names(assumptions[[scenario]]), function(id) {
    x <- assumptions[[scenario]][[id]]
    data.frame(scenario, field = id,
      value = as.character(jsonlite::toJSON(x$display$value, auto_unbox = TRUE)),
      source = paste(x$display$source, collapse = "; "))
  }))
}))
knitr::kable(rows)
```
